### 2020-11-18 doc/manuscript/Dow_manuscript.Rmd
add paragraph on data cleaning

(*The following seems out of date, right? Scripted cleaning is not described.*)
The dendroband data were manually cleaned by visual inspection. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. 3. Other - cases where data was clearly wrong but with unknown causes. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

@@

After model fitting, we used  a script to remove outliers....

@@

*(Numbers in bold will change)*

@@

~~For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest. Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.~~

~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest. Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.~~

>>>>

(*The following seems out of date, right? Scripted cleaning is not described.* -- *Manually cleaning was done for pre-analysis data. Added some text to clarify.*)
The dendroband data were manually cleaned by visual inspection before being used for modeling. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. 3. Other - cases where data was clearly wrong but with unknown causes. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

@@

After model fitting, we used a script to remove poorly fit models. Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the acceptable range we determined. Modeled tag years were removed if: 1. single day growth rates were higher than 2 standard deviations away from the mean for each wood-type (SCBI = 2, HF = 34) 2. DOY where peak growth rate occured was 2 standard deviations away, calculated for each wood-type in each year (aimed to remove models which predicted growth to peak during the winter; SCBI = 62, HF = 106) 3. tag years with small or negligible total growth (deltaDBH < 0.02 mm; SCBI = 0, HF = 66), 4. models which predicted total yearly growth to take longer than 365 days (SCBI = 181, HF = 199) 5. models with unexplained spikes (SCBI = 0, HF = 3). At Harvard Forest the tag years removed through this method was proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others *(SI figure? data_cleaning_figure_hf in manuscript folder)*. At SCBI a higher proportion of oak trees were removed *(data_cleaning_figure_scbi)*. The majority (White oak = 91, Red oak = 79) were removed in step 5 above, indicating that the LG5 model struggled to accurately model these these species.

@@

*(Numbers in bold will change, updated; 11/18. Should be final results)*

@@

For ring-porous trees, on average, 25% of annual growth was achieved on **May 4** ($DOY_{25}$=**123**) at SCBI and **May 15** ($DOY_{25}$=**132**) at Harvard FOrest, 50% on **June 2** ($DOY_{50}$=**152**) at SCBI and **June 9** ($DOY_{50}$=**159**) at Harvard Forest, and 75% on **June 30** ($DOY_{75}$=**180**) at SCBI and **July 6** ($DOY_{75}$=**186**) at Harvard Forest, for an average peak growing season length of **56** days at SCBI and **54** days at Harvard Forest. Maximum growth rate occurred *on $DOY_{50}$* at SCBI and *June 11* ($DOY_{max}$=**161**) at Harvard Forest and averaged .046 and .030 **mm day^-1^ (?)** respectively.

For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 4** ($DOY_{25}$=**154**) at SCBI and **June 14** ($DOY_{25}$=**164**) at Harvard Forest, 50% on **June 22** ($DOY_{50}$=**172**) at SCBI and **July 2** ($DOY_{50}$=**182**) at Harvard Forest, and 75% on **July 10** ($DOY_{75}$=**190**) and **July 19** ($DOY_{75}$=**199**) at Harvard forest, for an average peak growing season length of **36** days at SCBI and **35** days at Harvard Forest. Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 23** ($DOY_{max}$=**173**) at SCBI and **just one day after $DOY_{50}$** on **July 3** ($DOY_{max}$=**183**) at Harvard Forest and averaged .061 and .025 **mm day^-1^ (?)** respectively.
